---
title: "Capstone_Feb_2022"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import all libraries

```{r Library Import}
library(readr)
library(readxl)
library(skimr)
library(dplyr)
library(Metrics)
library(randomForest)
library(data.table)
library(caret)
library(ggplot2)

library(rpart)
library(rpart.plot)
```


Import and review data

test = flights_orig[c("Origin","Dist_from_DXB_NM")]
test$Dist_from_DXB_NM = as.numeric(test$Dist_from_DXB_NM)
unique(test[is.na(test$Dist_from_DXB_NM),])


```{r Import and Review Data}
flights_orig<- read_excel("DataDump_Full_v5.xlsx")

flights <- flights_orig[c("mins_flight120nmtoaldt",
                          "Airline_Categorisation",
                          "At120NM_Event",
                          "At120NM_Time",
                          "DC_120NMTime",
                          "eWTS_Trial_Live",
                          "Peak_Hours_120NMTime",
                          "rwy_config_at120NM",
                          "vis_cond_at120nm_mapped",
                          "Hour_120NM_Time",
                          "Wake_Vortex_Category",
                          "RECAT_Category",
                          "Category", 
                          "Engine_Type_Group",
                          "Congestion_Wake_Super", 
                          "Congestion_Wake_Heavy",
                          "Congestion_Wake_Medium",
                          "Congestion_Wake_Light",
                          "Congestion_Wake_Other",
                          "Congestion_Overall",
                          "Dist_from_DXB_NM",
                          "ALDT")]

#distance showing up as factor but should be numeric
flights$Dist_from_DXB_NM = as.numeric(flights$Dist_from_DXB_NM)

#remove entries where distance is null (<1% of dataset)
flights = flights[!is.na(flights$Dist_from_DXB_NM),]


#convert character variables to factor variables
flights[sapply(flights,is.character)]<-lapply(flights[sapply(flights, is.character)], as.factor)
flights$eWTS_Trial_Live<-as.factor(flights$eWTS_Trial_Live)



# Adjust to look at sample or all data; comment out at the end
# flights = head(flights, 50000)
```


Skim reviw
```{r Review}
#examine data
skim(flights)

```


Remove outliers
Note: we can either remove from both training and test, and call out in our test results (essentially saying x% is outside of the parameter of our model, and we expect x% of flights to be unpredictable in this way), or leave them in test results and also call out that we left all scenarios in

```{r Remove Outliers}
#Business intelligence says that the fastest a plane can travel 120nm is 5 mins. 
#remove all observations whose 120 mins flight is less than 5

# calculate percent of flights under 60 minutes
flights_under_60_percent<-nrow(subset(flights, flights$mins_flight120nmtoaldt<60))/length(flights_orig$mins_flight120nmtoaldt)

#we will never be able to train a model for extreme outliers so remove those as well
#flights_orig<-flights_orig[!(flights_orig$mins_flight120nmtoaldt <5), ] #why touch the original?
#flights_orig<-flights_orig[!(flights_orig$mins_flight120nmtoaldt >=60), ] #why touch the original?
flights<-flights[!(flights$mins_flight120nmtoaldt >=60), ]
flights<-flights[!(flights$mins_flight120nmtoaldt < 5), ]

```



```{r Day of the Week Feature}

# Add day of the week: want the day of the week to correspond to when at 120nm
flights$day_of_the_week = as.POSIXlt(flights$At120NM_Time)$wday
flights$day_of_the_week = as.character(flights$day_of_the_week)


# Add for visualization
# flights$day_and_hour = paste(flights$day_of_the_week, flights$Hour_120NM_Time)
```


```{r correlation}

# Check correlation between overall congestion and arrival time
cor(flights$Congestion_Overall, flights$mins_flight120nmtoaldt)

```


Additional data preparation for models (note that we prepare different datasets for different types of models: for example, XG Boost and random forest tree models; the only difference is that XG Boost requires categorical variables to be one-hot encoded while random forest models do not want one-hot encoding):
 - change all character variables to factor variables again
 - one-hot encode for regression and XG Boost models

```{r Data Prep}

#change all character variables to factor variables

flights[sapply(flights,is.character)]<-lapply(flights[sapply(flights, is.character)], as.factor)
flights$eWTS_Trial_Live<-as.factor(flights$eWTS_Trial_Live)
flights$Hour_120NM_Time<-as.factor(flights$Hour_120NM_Time)

#remove POSIIXct variables because they were only needed to produce derived variables & offer no predictive value
flights<-subset(flights, select = -c(At120NM_Time, ALDT))

#one-hot encode all categorical variables
hot_features<-c("Airline_Categorisation",
                "At120NM_Event", 
                "eWTS_Trial_Live",
                "Peak_Hours_120NMTime",
                "rwy_config_at120NM",
                "vis_cond_at120nm_mapped",
                "Wake_Vortex_Category", 
                "RECAT_Category",
                "Category",
                "Engine_Type_Group",
                "Hour_120NM_Time",
                "day_of_the_week")

dummies<- dummyVars( ~ Airline_Categorisation + 
                       At120NM_Event + 
                       eWTS_Trial_Live + 
                       Peak_Hours_120NMTime + 
                       rwy_config_at120NM + 
                       vis_cond_at120nm_mapped + 
                       Wake_Vortex_Category + 
                       RECAT_Category + 
                       Category + 
                       Engine_Type_Group + 
                       Hour_120NM_Time +
                       day_of_the_week, 
                       data = flights )

flight_cat<-as.data.frame(predict(dummies, newdata = flights))

#combine encoded variables with numeric variables
hot_flights<-cbind(flights[, -c(which(colnames(flights) %in% hot_features))],flight_cat)

#View(head(hot_flights))

```


```{r Training and Testing Splits based on Random Sample}
# Create test and training sets - regular
# Use 31 seed to stay consistent with Forrest

smp_size = floor(.75 * nrow(flights))
set.seed(31)
train_ind = sample(seq_len(nrow(flights)), size = smp_size)
flights_train = flights[train_ind, ]
flights_test = flights[-train_ind, ]

# Create test and training sets - one hot encoded
# Use 31 seed to stay consistent with Forrest

hot_flights_train = hot_flights[train_ind, ]
hot_flights_test = hot_flights[-train_ind, ]

```

```{r Evaluation Metrics}

#evaluate performance with MAE; how far on average is prediction from true value
MAE<-function(actual, predicted){
  mean(abs(actual-predicted))
}

#evaluate performance with percent of total predictions within target (3)
target<-3

Percent_within_tolerance = function(actual, predicted) {
  sum(abs(actual-predicted) <=target)/length(actual)
}

```




```{r Simple Congestion Regression}

congest_lm <- lm(mins_flight120nmtoaldt ~ Congestion_Overall, data=flights_train)
summary(congest_lm)

lm_prediction = predict(congest_lm, newdata = flights_test)

```


```{r Regression}

mutlivariable_lm <- lm(mins_flight120nmtoaldt ~  Congestion_Overall + Dist_from_DXB_NM + Airline_Categorisation.UAE + At120NM_Event.EES1.NAS + At120NM_Event.ENR1.NAS + At120NM_Event.EST1.NAS + `Peak_Hours_120NMTime.Off-Peak Hours` + eWTS_Trial_Live.1 + vis_cond_at120nm_mapped.Mediocre + vis_cond_at120nm_mapped.Poor + Wake_Vortex_Category.M + Wake_Vortex_Category.H + Wake_Vortex_Category.J + `rwy_config_at120NM.30R ARR 30L DEP` + `rwy_config_at120NM.Dual Runway 12` + `rwy_config_at120NM.Dual Runway 30` + `rwy_config_at120NM.Single Runway 12R` + `rwy_config_at120NM.Single Runway 30L` + day_of_the_week.0 + day_of_the_week.1 + day_of_the_week.2 + day_of_the_week.3 + day_of_the_week.4 + day_of_the_week.5, data=hot_flights_train)
summary(mutlivariable_lm)

lm_mv_prediction = predict(mutlivariable_lm, newdata = hot_flights_test)


```


```{r Decision Tree}
# training a model on the data using recursive partitioning 
# (basic numeric decision tree)

reg_tree_model<-rpart(mins_flight120nmtoaldt ~., data = flights_train, cp=0.009)
rpart.plot(reg_tree_model, digits = 1)
minute_predict<-predict(reg_tree_model, flights_test)
```


```{r Result Table 1 Data Frame}
# Create an data frame to store results

results <- data.frame(matrix(ncol = 6, nrow = 0))
names(results) <- c("Model", "Correlation", "MAE", "Relative Error", "RMSE", "% in Target")

#create vector of difference between predictions and actuals
avg_land<-mean(flights_train$mins_flight120nmtoaldt)
avg_prediction<-rep(avg_land, length(flights_test$mins_flight120nmtoaldt))

#the correlation between the predicted and actual quality values can gauge performance
#measure how strongly the predictions are related to true value; not a measure of how far off the predictions were from true values.

results[nrow(results) +1, ]<-list(Model = "Average of Actuals",
                                  Correlation = round(cor(avg_prediction,flights_test$mins_flight120nmtoaldt),1),
                                  MAE =round(MAE(avg_prediction, flights_test$mins_flight120nmtoaldt),1),
                                  MAPE =round(mape(avg_prediction, flights_test$mins_flight120nmtoaldt)*100,1),
                                  RMSE =round(rmse(avg_prediction, flights_test$mins_flight120nmtoaldt),1),
                                  Percent_in_tolerance = round(Percent_within_tolerance(flights_test$mins_flight120nmtoaldt, avg_prediction)*100,1))

results[nrow(results) +1, ]<-list(Model = "Linear Regression - Pure Congestion",
                                  Correlation = round(cor(lm_prediction,hot_flights_test$mins_flight120nmtoaldt),1),
                                  MAE =round(MAE(lm_prediction, hot_flights_test$mins_flight120nmtoaldt),1),
                                  MAPE =round(mape(lm_prediction, hot_flights_test$mins_flight120nmtoaldt)*100,1),
                                  RMSE =round(rmse(lm_prediction, hot_flights_test$mins_flight120nmtoaldt),1),
                                  Percent_within_tolerance = round(Percent_within_tolerance(hot_flights_test$mins_flight120nmtoaldt, lm_prediction)*100,1))

                                
results[nrow(results) +1, ]<-list(Model = "Linear Regression - Multi-Variable",
                                  Correlation = round(cor(lm_mv_prediction,hot_flights_test$mins_flight120nmtoaldt),1),
                                  MAE =round(MAE(lm_mv_prediction, hot_flights_test$mins_flight120nmtoaldt),1),
                                  MAPE =round(mape(lm_mv_prediction, hot_flights_test$mins_flight120nmtoaldt)*100,1),
                                  RMSE =round(rmse(lm_mv_prediction, hot_flights_test$mins_flight120nmtoaldt),1),
                                  Percent_in_tolerance = round(Percent_within_tolerance(hot_flights_test$mins_flight120nmtoaldt, lm_mv_prediction)*100,1))
                              
                                  
        
results[nrow(results) +1, ]<-list(Model = "Decision Tree",
                                  Correlation = round(cor(minute_predict,flights_test$mins_flight120nmtoaldt),1),
                                  MAE =round(MAE(minute_predict,flights_test$mins_flight120nmtoaldt),1),
                                  MAPE =round(mape(minute_predict,flights_test$mins_flight120nmtoaldt)*100,1),
                                  RMSE =round(rmse(minute_predict,flights_test$mins_flight120nmtoaldt),1),
                                  Percent_in_tolerance = round(Percent_within_tolerance(flights_test$mins_flight120nmtoaldt, minute_predict)*100,1))

results
```
